---
title: "Coursera Getting Human Activity Recognition DataSet"
author: "Jeremy Erickson"
date: "November 11, 2018"
output: html_document
---

## Background Information:


### **Information on the dataset**:

\"The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data.  

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.\"   

\"For each record it is provided:
  - Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
  - Triaxial Angular velocity from the gyroscope. 
  - A 561-feature vector with time and frequency domain variables. 
  - Its activity label. 
  - An identifier of the subject who carried out the experiment.

**The dataset includes the following files:**

  - README.txt
  - features_info.txt: Shows information about the variables used on the feature vector.
  - features.txt: List of all features.
  - activity_labels.txt: Links the class labels with their activity name.
  - train/X_train.txt: Training set.
  - train/y_train.txt: Training labels.
  - test/X_test.txt: Test set.
  - test/y_test.txt: Test labels.

**The following files are available for the train and test data. Their descriptions are equivalent.** 

  - train/subject_train.txt: Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 
  - train/Inertial Signals/total_acc_x_train.txt: The acceleration signal from the smartphone accelerometer X axis in standard gravity units 'g'.  
    Every row shows a 128 element vector. The same description applies for the total_acc_x_train.txt and total_acc_z_train.txt files for the Y and Z         axis. 
  - train/Inertial Signals/body_acc_x_train.txt: The body acceleration signal obtained by subtracting the gravity from the total acceleration. 
  - train/Inertial Signals/body_gyro_x_train.txt': The angular velocity vector measured by the gyroscope for each window sample. The units are               radians/second. \"

\"The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

**These signals were used to estimate variables of the feature vector for each pattern:**  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

  - tBodyAcc-XYZ
  - tGravityAcc-XYZ
  - tBodyAccJerk-XYZ
  - tBodyGyro-XYZ
  - tBodyGyroJerk-XYZ
  - tBodyAccMag
  - tGravityAccMag
  - tBodyAccJerkMag
  - tBodyGyroMag
  - tBodyGyroJerkMag
  - fBodyAcc-XYZ
  - fBodyAccJerk-XYZ
  - fBodyGyro-XYZ
  - fBodyAccMag
  - fBodyAccJerkMag
  - fBodyGyroMag
  - fBodyGyroJerkMag

**The set of variables that were estimated from these signals are:**

  - mean(): Mean value
  - std(): Standard deviation
  - mad(): Median absolute deviation 
  - max(): Largest value in array
  - min(): Smallest value in array
  - sma(): Signal magnitude area
  - energy(): Energy measure. Sum of the squares divided by the number of values. 
  - iqr(): Interquartile range 
  - entropy(): Signal entropy
  - arCoeff(): Autorregresion coefficients with Burg order equal to 4
  - correlation(): correlation coefficient between two signals
  - maxInds(): index of the frequency component with largest magnitude
  - meanFreq(): Weighted average of the frequency components to obtain a mean frequency
  - skewness(): skewness of the frequency domain signal 
  - kurtosis(): kurtosis of the frequency domain signal 
  - bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
  - angle(): Angle between to vectors.

**Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:**

  - gravityMean
  - tBodyAccMean
  - tBodyAccJerkMean
  - tBodyGyroMean
  - tBodyGyroJerkMean\"


**The full dataset can be download from the following site:**

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones


**Reference and copied from:**

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. 
Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine.  
International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012   



### **Getting and Cleaning Data Coursera course:**

The final project of the Getting and Cleaning data course uses a portion of the orginal dataset. 

**The course site is listed below:**

https://www.coursera.org/learn/data-cleaning/peer/FIZtT/getting-and-cleaning-data-course-project

The goal of this project is to create a summarized report. The report will include 66 out of the total 561 features captured. These features were classified as either being a mean or a standard deviation. These features are listed below:

  - tBodyAcc-mean()-X
  - tBodyGyro-mean()-X
  - tBodyGyro-mean()-Y
  - tBodyGyro-mean()-Z
  - tBodyGyro-std()-X
  - tBodyGyro-std()-Y
  - tBodyGyro-std()-Z
  - tBodyGyroJerk-mean()-X
  - tBodyGyroJerk-mean()-Y
  - tBodyGyroJerk-mean()-Z
  - tBodyGyroJerk-std()-X
  - tBodyGyroJerk-std()-Y
  - tBodyGyroJerk-std()-Z
  - tBodyAcc-mean()-Y
  - tBodyAccMag-mean()
  - tBodyAccMag-std()
  - tGravityAccMag-mean()
  - tGravityAccMag-std()
  - tBodyAccJerkMag-mean()
  - tBodyAccJerkMag-std()
  - tBodyGyroMag-mean()
  - tBodyGyroMag-std()
  - tBodyGyroJerkMag-mean()
  - tBodyGyroJerkMag-std()
  - fBodyAcc-mean()-X
  - fBodyAcc-mean()-Y
  - fBodyAcc-mean()-Z
  - fBodyAcc-std()-X
  - fBodyAcc-std()-Y
  - fBodyAcc-std()-Z
  - tBodyAcc-mean()-Z
  - fBodyAccJerk-mean()-X
  - fBodyAccJerk-mean()-Y
  - fBodyAccJerk-mean()-Z
  - fBodyAccJerk-std()-X
  - fBodyAccJerk-std()-Y
  - fBodyAccJerk-std()-Z
  - tBodyAcc-std()-X
  - tGravityAcc-mean()-X
  - tGravityAcc-mean()-Y
  - fBodyGyro-mean()-X
  - fBodyGyro-mean()-Y
  - fBodyGyro-mean()-Z
  - fBodyGyro-std()-X
  - fBodyGyro-std()-Y
  - fBodyGyro-std()-Z
  - tGravityAcc-mean()-Z
  - tGravityAcc-std()-X
  - tGravityAcc-std()-Y
  - tGravityAcc-std()-Z
  - tBodyAcc-std()-Y
  - fBodyAccMag-mean()
  - fBodyAccMag-std()
  - fBodyBodyAccJerkMag-mean()
  - fBodyBodyAccJerkMag-std()
  - fBodyBodyGyroMag-mean()
  - fBodyBodyGyroMag-std()
  - fBodyBodyGyroJerkMag-mean()
  - fBodyBodyGyroJerkMag-std()
  - tBodyAcc-std()-Z
  - tBodyAccJerk-mean()-X
  - tBodyAccJerk-mean()-Y
  - tBodyAccJerk-mean()-Z
  - tBodyAccJerk-std()-X
  - tBodyAccJerk-std()-Y
  - tBodyAccJerk-std()-Z

These features were captured multiple times for 30 subjects in performing 6 diffent activities. The tidy dataset provided in this repository is the mean of all 66 features grouped by subject by activity creating a data table of 180 rows by 68 collumns.   

## Complete File Listing:

```{r echo=FALSE, message=FALSE}
library(dplyr)
library(knitr)
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileUrl, destfile="./data/dataset.zip")
unzip(zipfile="./data/dataset.zip")
files<-list.files("./data", recursive=TRUE)
files
```

## File Description:
#### Only the following files will be used:  
**1. activity_labels.txt** - contains the names of the activities.  
      - There are 6 activities   
**2. features.txt** - contains the names of the features captured.  
      - There are 651 features captured.  
**3. subject_test.txt** - the test subjects maps the y_test data  
      - There are 10 test subjects  
**4. X_test.txt** - multiple test feature measurements for the 10 subjects     
**5. y_test.txt** - the test subjects rows .. need to add this column to the data  
**6. subject_train.txt** - the training data subjects maps the y_train data  
      - There are 20 training subjects  
**7. X_train.txt** - multiple training feature measurements for the 20 subjects    
**8. y_train.txt** - the train subjects .. need to add this column to the data    

## Algorithm Recipe:

### Goal:
The goal of this project is to create a tidy dataset that will contain the average values of 30 subjects across 6 different activities for only the features that are classified as mean or standard deviation values. This dataset needs to be properly labeled. 

### Step 1: Reading in all of the data:  
#####Code:  
 **# read the activity labels into R.**  
activities <- read.table("./data/UCI HAR Dataset/activity_labels.txt")   
 **# read in the 561 feature labels into R**  
features <- read.table("./data/UCI HAR Dataset/features.txt")   
 **# read the test data into R**  
testYdata <- read.table("./data/UCI HAR Dataset/test/y_test.txt")  
testXdata <- read.table("./data/UCI HAR Dataset/test/X_test.txt")  
testSubjectdata <- read.table("./data/UCI HAR Dataset/test/subject_test.txt")  
 **#read the training data into R**  
trainYdata <- read.table("./data/UCI HAR Dataset/train/y_train.txt")  
trainXdata <- read.table("./data/UCI HAR Dataset/train/X_train.txt")  
trainSubjectdata <- read.table("./data/UCI HAR Dataset/train/subject_train.txt")  
 **#quick check on the dimensions of the data just to make sure**  
 **# the column numbers are the same for the data**  
dim(testXdata)  
 **#[1] 2947  561**  
dim(trainXdata)  
 **#[1] 7352  561**  
 **# quick check to make sure the row counts are the same for the respective X data set.**  
dim(testYdata)  
 **#[1] 2947    1**  
dim(trainYdata)  
 **#[1] 7352    1**  
dim(testSubjectdata)  
 **#[1] 2947    1**   
dim(trainSubjectdata)  
 **#[1] 7352    1**

```{r echo=FALSE, message=FALSE}
activities <- read.table("./data/UCI HAR Dataset/activity_labels.txt")   
features <- read.table("./data/UCI HAR Dataset/features.txt")   
testYdata <- read.table("./data/UCI HAR Dataset/test/y_test.txt")  
testXdata <- read.table("./data/UCI HAR Dataset/test/X_test.txt")  
testSubjectdata <- read.table("./data/UCI HAR Dataset/test/subject_test.txt")  
trainYdata <- read.table("./data/UCI HAR Dataset/train/y_train.txt")  
trainXdata <- read.table("./data/UCI HAR Dataset/train/X_train.txt")  
trainSubjectdata <- read.table("./data/UCI HAR Dataset/train/subject_train.txt")  
```

### Step 2: Creating the combined dataset:
#####Code: Combining the test and training dataset.
 **# Now we are combining the test and training data**  
xcombine <- rbind(testXdata,trainXdata)  
ycombine <- rbind(testYdata,trainYdata)  
subjectcombine <- rbind(testSubjectdata,trainSubjectdata)  
 **# Adding the column headings (features) to the measurement file.**  
names(xcombine)<- features$V2  
 **# Now we only want to keep the measurements that are labelled mean or std**  
keepfeatures <- grep("-mean\\(\\)|-std\\(\\)", features\$V2)   
xselected <-  xcombine[, keepfeatures]  
 **# Adding the activity label**  
names(ycombine) <- c("activity")  
 **# Adding the subject label**  
names(subjectcombine) <- c("subject")  
 **# Creating the first complete dataset**  
Data <- cbind(subjectcombine,ycombine,xselected)  

```{r echo=FALSE, message=FALSE}
xcombine <- rbind(testXdata,trainXdata)  
ycombine <- rbind(testYdata,trainYdata)  
subjectcombine <- rbind(testSubjectdata,trainSubjectdata)  
names(xcombine)<- features$V2  
keepfeatures <- grep("-mean\\(\\)|-std\\(\\)", features$V2)   
xselected <-  xcombine[, keepfeatures]  
names(ycombine) <- c("activity")  
names(subjectcombine) <- c("subject")  
Data <- cbind(subjectcombine,ycombine,xselected)  
```

### Step 3: Cleaning up the dataset:
#####Code:  This code cleans the labels and converts the activity IDs to Labels

 **# removes the -, the (, and the ) character from the label**  
names(Data) <- gsub("\\\\\-\|\\\\(\|\\\\)","", names(Data))  
 **# substitutes the beginning t with time**  
names(Data) <- gsub("\^t","time", names(Data))  
 **# substitutes the beginning f with frequency**  
names(Data) <- gsub("\^f","frequency", names(Data))  
 **# substitutes Acc with accelerometer**  
names(Data)<-gsub("Acc", "accelerometer", names(Data))  
 **# substitutes Gyro with gyroscope**  
names(Data)<-gsub("Gyro", "gyroscope", names(Data))  
 **# substitutes Mag with magnitude**  
names(Data)<-gsub("Mag", "magnitude", names(Data))  
 **# substitutes BodyBody with just body**  
names(Data)<-gsub("BodyBody", "body", names(Data))  
 **# converts all letters to lower case**  
names(Data)<-tolower(names(Data))  
 **# Now we will change the activity numbers to the proper labels.**   
Data[, 2] <- activities[Data[,2], 2]  
 **# We also want to change the subject from a number class to a factor**  
Data[, 1] <- as.factor(Data[,1])  

      Orginal                       TidyName                                     
    [1,] "subject"                     "subject"                                    
    [2,] "activity"                    "activity"                                   
    [3,] "tBodyAcc-mean()-X"           "timebodyaccelerometermeanx"                 
    [4,] "tBodyAcc-mean()-Y"           "timebodyaccelerometermeany"                 
    [5,] "tBodyAcc-mean()-Z"           "timebodyaccelerometermeanz"                 
    [6,] "tBodyAcc-std()-X"            "timebodyaccelerometerstdx"                  
    [7,] "tBodyAcc-std()-Y"            "timebodyaccelerometerstdy"                  
    [8,] "tBodyAcc-std()-Z"            "timebodyaccelerometerstdz"                  
    [9,] "tGravityAcc-mean()-X"        "timegravityaccelerometermeanx"              
    [10,] "tGravityAcc-mean()-Y"        "timegravityaccelerometermeany"              
    [11,] "tGravityAcc-mean()-Z"        "timegravityaccelerometermeanz"              
    [12,] "tGravityAcc-std()-X"         "timegravityaccelerometerstdx"               
    [13,] "tGravityAcc-std()-Y"         "timegravityaccelerometerstdy"               
    [14,] "tGravityAcc-std()-Z"         "timegravityaccelerometerstdz"               
    [15,] "tBodyAccJerk-mean()-X"       "timebodyaccelerometerjerkmeanx"             
    [16,] "tBodyAccJerk-mean()-Y"       "timebodyaccelerometerjerkmeany"             
    [17,] "tBodyAccJerk-mean()-Z"       "timebodyaccelerometerjerkmeanz"             
    [18,] "tBodyAccJerk-std()-X"        "timebodyaccelerometerjerkstdx"              
    [19,] "tBodyAccJerk-std()-Y"        "timebodyaccelerometerjerkstdy"              
    [20,] "tBodyAccJerk-std()-Z"        "timebodyaccelerometerjerkstdz"              
    [21,] "tBodyGyro-mean()-X"          "timebodygyroscopemeanx"                     
    [22,] "tBodyGyro-mean()-Y"          "timebodygyroscopemeany"                     
    [23,] "tBodyGyro-mean()-Z"          "timebodygyroscopemeanz"                     
    [24,] "tBodyGyro-std()-X"           "timebodygyroscopestdx"                      
    [25,] "tBodyGyro-std()-Y"           "timebodygyroscopestdy"                      
    [26,] "tBodyGyro-std()-Z"           "timebodygyroscopestdz"                      
    [27,] "tBodyGyroJerk-mean()-X"      "timebodygyroscopejerkmeanx"                 
    [28,] "tBodyGyroJerk-mean()-Y"      "timebodygyroscopejerkmeany"                 
    [29,] "tBodyGyroJerk-mean()-Z"      "timebodygyroscopejerkmeanz"                 
    [30,] "tBodyGyroJerk-std()-X"       "timebodygyroscopejerkstdx"                  
    [31,] "tBodyGyroJerk-std()-Y"       "timebodygyroscopejerkstdy"                  
    [32,] "tBodyGyroJerk-std()-Z"       "timebodygyroscopejerkstdz"                  
    [33,] "tBodyAccMag-mean()"          "timebodyaccelerometermagnitudemean"         
    [34,] "tBodyAccMag-std()"           "timebodyaccelerometermagnitudestd"          
    [35,] "tGravityAccMag-mean()"       "timegravityaccelerometermagnitudemean"      
    [36,] "tGravityAccMag-std()"        "timegravityaccelerometermagnitudestd"       
    [37,] "tBodyAccJerkMag-mean()"      "timebodyaccelerometerjerkmagnitudemean"     
    [38,] "tBodyAccJerkMag-std()"       "timebodyaccelerometerjerkmagnitudestd"      
    [39,] "tBodyGyroMag-mean()"         "timebodygyroscopemagnitudemean"             
    [40,] "tBodyGyroMag-std()"          "timebodygyroscopemagnitudestd"              
    [41,] "tBodyGyroJerkMag-mean()"     "timebodygyroscopejerkmagnitudemean"         
    [42,] "tBodyGyroJerkMag-std()"      "timebodygyroscopejerkmagnitudestd"          
    [43,] "fBodyAcc-mean()-X"           "frequencybodyaccelerometermeanx"            
    [44,] "fBodyAcc-mean()-Y"           "frequencybodyaccelerometermeany"            
    [45,] "fBodyAcc-mean()-Z"           "frequencybodyaccelerometermeanz"            
    [46,] "fBodyAcc-std()-X"            "frequencybodyaccelerometerstdx"             
    [47,] "fBodyAcc-std()-Y"            "frequencybodyaccelerometerstdy"             
    [48,] "fBodyAcc-std()-Z"            "frequencybodyaccelerometerstdz"             
    [49,] "fBodyAccJerk-mean()-X"       "frequencybodyaccelerometerjerkmeanx"        
    [50,] "fBodyAccJerk-mean()-Y"       "frequencybodyaccelerometerjerkmeany"        
    [51,] "fBodyAccJerk-mean()-Z"       "frequencybodyaccelerometerjerkmeanz"        
    [52,] "fBodyAccJerk-std()-X"        "frequencybodyaccelerometerjerkstdx"         
    [53,] "fBodyAccJerk-std()-Y"        "frequencybodyaccelerometerjerkstdy"         
    [54,] "fBodyAccJerk-std()-Z"        "frequencybodyaccelerometerjerkstdz"         
    [55,] "fBodyGyro-mean()-X"          "frequencybodygyroscopemeanx"                
    [56,] "fBodyGyro-mean()-Y"          "frequencybodygyroscopemeany"                
    [57,] "fBodyGyro-mean()-Z"          "frequencybodygyroscopemeanz"                
    [58,] "fBodyGyro-std()-X"           "frequencybodygyroscopestdx"                 
    [59,] "fBodyGyro-std()-Y"           "frequencybodygyroscopestdy"                 
    [60,] "fBodyGyro-std()-Z"           "frequencybodygyroscopestdz"                 
    [61,] "fBodyAccMag-mean()"          "frequencybodyaccelerometermagnitudemean"    
    [62,] "fBodyAccMag-std()"           "frequencybodyaccelerometermagnitudestd"     
    [63,] "fBodyBodyAccJerkMag-mean()"  "frequencybodyaccelerometerjerkmagnitudemean"
    [64,] "fBodyBodyAccJerkMag-std()"   "frequencybodyaccelerometerjerkmagnitudestd" 
    [65,] "fBodyBodyGyroMag-mean()"     "frequencybodygyroscopemagnitudemean"        
    [66,] "fBodyBodyGyroMag-std()"      "frequencybodygyroscopemagnitudestd"         
    [67,] "fBodyBodyGyroJerkMag-mean()" "frequencybodygyroscopejerkmagnitudemean"    
    [68,] "fBodyBodyGyroJerkMag-std()"  "frequencybodygyroscopejerkmagnitudestd"  




```{r echo=FALSE, message=FALSE}
DataA <- cbind(subjectcombine,ycombine,xselected) 
names(Data) <- gsub("\\-|\\(|\\)","", names(Data))  
names(Data) <- gsub("^t","time", names(Data))  
names(Data) <- gsub("^f","frequency", names(Data))  
names(Data)<-gsub("Acc", "accelerometer", names(Data))  
names(Data)<-gsub("Gyro", "gyroscope", names(Data))  
names(Data)<-gsub("Mag", "magnitude", names(Data))  
names(Data)<-gsub("BodyBody", "body", names(Data))  
names(Data)<-tolower(names(Data))  
Data[, 2] <- activities[Data[,2], 2]  
Data[, 1] <- as.factor(Data[,1])  

```

### Step 4: Creating the tidy dataset:  
#####Code: Computing the mean for each feature grouped by subject, activity  

tidyData <- group_by(Data,activity,subject)  
 **# Taking the mean of the dataset**  
meantidyData <- summarize_all(tidyData, mean, na.rm=TRUE)  
 **# Exporting the tidy dataset to a file**  
write.table(meantidyData, "fitness_tidy_dataset.txt", row.names = FALSE) 

```{r echo=FALSE, message=FALSE}
tidyData <- group_by(Data,activity,subject)  
meantidyData <- summarize_all(tidyData, mean, na.rm=TRUE)  
write.table(meantidyData, "fitness_tidy_dataset.txt", row.names = FALSE)
```
## Requirements and Code:

This code is a R script. The additional packages required for running this code are dplyr and knitr.

**This code was ran with the following conditions:**

  - R version 3.5.1 (2018-07-02)
  - Platform: x86_64-w64-mingw32/x64 (64-bit)
  - Running under: Windows >= 8 x64 (build 9200)
  - knitr_1.20   
  - dplyr_0.7.7

